---
title: "Additive P-splines in PROregTMB"
author: "Dae-Jin Lee"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Additive P-splines in PROregTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.8,
  dpi = 120
)
```

# Model

Additive BB regression on the logit scale with Eilers (1999) mixed
reparameterisation (same engine as BB-GAM / Report 4):

\[
\operatorname{logit}(p_i)
= x_i^\top\beta + \sum_{j=1}^{J} f_j(x_{ij}),\qquad
f_j = X_{\mathrm{null},j}\beta_{\mathrm{null},j} + Z_j s_j,
\quad s_j \sim N(0,\sigma_{s_j}^2 I).
\]

For each smooth, \(Z_j = B D'(DD')^{-1}\) from the Eilers–Marx B-spline
basis and difference penalty of order `pord`, residualised against the
fixed design so the polynomial null space lives in \(X\).

Wiggly coefficients \(s_j\) are integrated out by Laplace approximation;
smooth SDs \(\sigma_{s_j}\) are estimated jointly (`smooth_sd` on the fit;
`lambda = 1/smooth_sd^2` is kept as an alias).

Notation follows Eilers: `ndx` (intervals), `pord` (difference order),
`bdeg` (B-spline degree).

```{r lib}
library(PROregTMB)
```

# Formula API

```{r api-demo, eval=FALSE}
BBreg(y ~ z + s(age, ndx = 12, pord = 2) + s(bmi, ndx = 10),
      m = m, data = dat)

BBmm(y ~ s(time, ndx = 10), random = ~ (1 | id),
     m = m, data = dat)
```

Defaults: `ndx = 10`, `pord = 2`, `bdeg = 3`.

# Cross-sectional additive example

Two centered smooths on independent covariates:

```{r sim-add}
set.seed(11)
n <- 300
m <- 10L
x1 <- runif(n)
x2 <- runif(n)
f1_true <- sin(2 * pi * x1)
f2_true <- exp(2 * x2) - (exp(2) - 1) / 2
eta <- -0.4 + f1_true + f2_true
y <- rBB(n, m, 1 / (1 + exp(-eta)), phi = 0.2)
dat <- data.frame(y, x1, x2)

fit <- BBreg(
  y ~ s(x1, ndx = 10, pord = 2) + s(x2, ndx = 10, pord = 2),
  m = m,
  data = dat
)
fit$conv
fit$lambda
```

Check the sum-to-zero constraints and recover component functions:

```{r checks}
vapply(fit$fhat, sum, numeric(1))
cor(fit$fhat[["s(x1)"]], f1_true - mean(f1_true))
cor(fit$fhat[["s(x2)"]], f2_true - mean(f2_true))
```

```{r plot-add, fig.cap="Estimated smooths with 95% pointwise bands vs centered truth (dashed)."}
truths <- list(f1_true - mean(f1_true), f2_true - mean(f2_true))
xs <- list(x1, x2)
labs <- names(fit$fhat)
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
for (j in 1:2) {
  g <- seq(min(xs[[j]]), max(xs[[j]]), length.out = 120)
  pr <- predict_smooth(fit, which = j, x = g, method = "marrawood")
  o <- order(xs[[j]])
  ylim <- range(pr$lwr, pr$upr, truths[[j]], finite = TRUE)
  plot(NA, xlim = range(g), ylim = ylim,
       xlab = paste0("x", j), ylab = paste0("f", j),
       main = labs[j])
  polygon(c(pr$x, rev(pr$x)), c(pr$lwr, rev(pr$upr)),
          col = adjustcolor("darkorange2", 0.30), border = NA)
  lines(xs[[j]][o], truths[[j]][o], lty = 2, lwd = 1.5)
  lines(pr$x[order(pr$x)], pr$fit[order(pr$x)], col = "darkorange2", lwd = 2)
  abline(h = 0, col = "grey70")
}
par(op)

# Helper: plot_smooth(fit)  # same bands
```

Pointwise bands use the Bayesian / Wahba–Silverman covariance of the
P-spline coefficients from the Laplace posterior
(\eqn{\hat f\pm z\sqrt{b^\top V_\gamma b}}), following Marra & Wood (2012;
see References). For `BBmm`, \(V_\gamma\) is the **marginal** covariance
of the spline block after integrating subject random effects.

# Mixed model with a smooth (`BBmm`)

```{r mm-smooth}
set.seed(12)
n_g <- 30
n_per <- 5
m <- 8L
id <- factor(rep(seq_len(n_g), each = n_per))
x <- runif(n_g * n_per)
u <- rnorm(n_g, 0, 0.5)
eta <- 0.2 + sin(2 * pi * x) + u[id]
y <- rBB(length(x), m, 1 / (1 + exp(-eta)), phi = 0.15)
dat_mm <- data.frame(y, x, id)

fit_mm <- BBmm(
  y ~ s(x, ndx = 10, pord = 2),
  random = ~ (1 | id),
  m = m,
  data = dat_mm
)
fit_mm$lambda
fit_mm$sigma
sum(fit_mm$fhat[[1]])
```

```{r plot-mm, fig.cap="BBmm smooth with 95% pointwise band vs centered truth (dashed)."}
g <- seq(0, 1, length.out = 100)
pr <- predict_smooth(fit_mm, which = 1, x = g, method = "marrawood")
ylim <- range(pr$lwr, pr$upr, sin(2 * pi * g) - mean(sin(2 * pi * x)),
              finite = TRUE)
plot(NA, xlim = range(g), ylim = ylim,
     xlab = "x", ylab = "f(x)", main = "BBmm + s(x)")
polygon(c(pr$x, rev(pr$x)), c(pr$lwr, rev(pr$upr)),
        col = adjustcolor("darkorange2", 0.30), border = NA)
lines(g, sin(2 * pi * g) - mean(sin(2 * pi * x)), lty = 2)
lines(pr$x, pr$fit, col = "darkorange2", lwd = 2)
abline(h = 0, col = "grey70")
legend("topright", c("true (centered)", "estimate", "95% band"),
       lty = c(2, 1, NA), lwd = c(1, 2, NA),
       fill = c(NA, NA, adjustcolor("darkorange2", 0.30)),
       border = c(NA, NA, NA),
       col = c("black", "darkorange2", NA), bty = "n")
```

# Helpers

| Object / function | Role |
|-------------------|------|
| `s(x, ndx, pord, bdeg)` | Smooth term in a formula |
| `bbase()` / `penalty_diff()` | Basis helpers (public) |
| `fit$smooth_sd` | Estimated smooth RE SDs \(\sigma_s\) |
| `fit$lambda` | Alias \(1/\sigma_s^2\) |
| `fit$fhat` | Fitted smooths at observed design |
| `fit$s` | Wiggly coefficients \(s\) |
| `predict_smooth(..., method="marrawood")` | Pointwise bands (Marra & Wood, 2012) |
| `plot_smooth()` | Plot estimate + shaded band |

Multivariate `cbind()` / `dim=` APIs accept **shared** `s()` terms, or
**domain-specific** `s(x, by = dim)` with long data + `dim =`. See the mixed
vignette and `scripts/fit_COPD_multi_pspline.R`.

# References

Eilers, P. H. C. (1999). Discussion of Currie and Durban.
*Journal of the Royal Statistical Society: Series C*, **48**, 269–270.

Marra, G. and Wood, S. N. (2012). Coverage properties of confidence intervals
for generalized additive model components.
*Scandinavian Journal of Statistics*, **39**(1), 53–74.
doi:[10.1111/j.1467-9469.2011.00760.x](https://doi.org/10.1111/j.1467-9469.2011.00760.x)
