---
title: "PROregTMB mixed and multivariate models"
author: "Dae-Jin Lee"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PROregTMB mixed and multivariate models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 4,
  dpi = 120
)
```

# Canonical mixed model

**PROregTMB** uses a shared-latent formulation for univariate and multivariate
BB mixed models:

\[
\operatorname{logit}(p_{ij}^{(\ell)})
= x_{ij}^{(\ell)\top}\beta^{(\ell)} + z_{ij}^{\top} a_i,\qquad
a_i\sim N(0,G),
\]

estimated by maximizing the Laplace approximation of the marginal likelihood
(TMB). Here \(\ell\) is the outcome dimension (\(L=1\) univariate), \(j\) indexes
visits (\(n_i=1\) for cross-sectional data), and
\(z_{ij}=(1)\) or \((1,t_{ij})^\top\).

```{r lib}
library(PROregTMB)
```

# Random intercept

```{r ri}
set.seed(42)
n_g <- 30
n_per <- 6
m <- 10L
x <- rnorm(n_g * n_per)
id <- factor(rep(seq_len(n_g), each = n_per))
u <- rnorm(n_g, 0, 0.8)
eta <- 0.5 - 0.4 * x + u[id]
y <- rBB(length(x), m, 1 / (1 + exp(-eta)), phi = 0.15)
dat <- data.frame(y, x, id)

fit_ri <- BBmm(y ~ x, random = ~ (1 | id), m = m, data = dat)
summary(fit_ri)
```

Primary fields:

```{r ri-fields}
fit_ri$beta
fit_ri$phi
fit_ri$sigma
head(fit_ri$u)
fit_ri$model
```

Legacy PROreg aliases (`$fixed.coef`, `$sigma.coef`, `$random.coef`,
`$phi.coef`) are still attached.

Crossed factors: `random = ~ (1 | site) + (1 | doc)` (or legacy
`random.formula = ~ site + doc`).

# Random intercept + random slope

This example uses 80 subjects with eight visits, centered time, and a
moderate random-slope SD (0.5) with dispersion 0.1 to make both random
effects identifiable in the simulated sample. Smaller samples with weak
random slopes can produce nearly singular covariance estimates.

```{r ri-rs}
set.seed(43)
n_g <- 80
n_per <- 8
m <- 10L
id <- factor(rep(seq_len(n_g), each = n_per))
time <- rep(seq(-1, 1, length.out = n_per), times = n_g)
G <- matrix(c(0.7^2, 0.3 * 0.7 * 0.5,
              0.3 * 0.7 * 0.5, 0.5^2), 2, 2)
a <- matrix(rnorm(n_g * 2), n_g, 2) %*% chol(G)
eta <- -0.2 + 0.15 * time + a[id, 1] + a[id, 2] * time
y <- rBB(length(time), m, 1 / (1 + exp(-eta)), phi = 0.1)
dat_rs <- data.frame(y, time, id)

fit_rs <- BBmm(
  y ~ time,
  random = ~ (1 + time | id),
  corr = "unstructured",
  m = m,
  data = dat_rs
)
fit_rs$sigma
fit_rs$Sigma$id
fit_rs$Corr$id
```

Check the optimizer status, gradient and Hessian together before using
the standard errors. A small gradient alone does not guarantee a valid
covariance estimate.

```{r ri-rs-diagnostics}
fit_rs$opt$convergence
fit_rs$max_grad
fit_rs$sdreport$pdHess
summary(fit_rs$sdreport, "report")
```

Use `corr = "diag"` for independent RI and RS. Aliases for unstructured:
`"us"`, `"cor"`, `"correlated"`.

# Multivariate shared latent

## Wide / cross-sectional (`cbind`)

```{r multi-wide}
set.seed(44)
n <- 80
id <- factor(seq_len(n))
x <- rnorm(n)
a <- rnorm(n, 0, 0.6)
p1 <- 1 / (1 + exp(-(0.2 - 0.3 * x + a)))
p2 <- 1 / (1 + exp(-(-0.1 + 0.4 * x + a)))
dat_w <- data.frame(
  id = id,
  x = x,
  y1 = rBB(n, 20L, p1, 0.15),
  y2 = rBB(n, 10L, p2, 0.2)
)

fit_w <- BBmm(
  cbind(y1, y2) ~ x,
  random = ~ (1 | id),
  m = c(20, 10),
  data = dat_w
)
fit_w$beta
fit_w$phi
fit_w$sigma
```

Dimension-specific fixed effects are named with the outcome prefix
(e.g. `y1.(Intercept)`, `y2.x`). The latent \(a_i\) is **shared**.

## Long format (`dim =`)

```{r multi-long}
set.seed(45)
n_g <- 50
id <- factor(rep(seq_len(n_g), each = 2))
domain <- factor(rep(c("SF36", "ACT"), times = n_g))
x <- rnorm(n_g * 2)
a <- rnorm(n_g, 0, 0.5)[as.integer(id)]
m_vec <- ifelse(domain == "SF36", 20L, 10L)
eta <- ifelse(domain == "SF36", 0.1, -0.2) + 0.25 * x + a
y <- rBB(length(x), m_vec, 1 / (1 + exp(-eta)), phi = 0.18)
long <- data.frame(y, x, id, domain, m = m_vec)

fit_l <- BBmm(
  y ~ x,
  random = ~ (1 | id),
  dim = "domain",
  m = "m",
  data = long
)
fit_l$beta
fit_l$phi
```

## M2 / M3: domain-specific random effects

Long format only. Indicator RE per domain for each subject
\(\mathbf{u}_i\sim N(0,\Sigma)\):

```{r multi-M3, eval=FALSE}
# M3: correlated domain RE (BB-GAM primary model)
BBmm(
  y ~ x,
  random = ~ (0 + domain | id),
  dim = "domain",
  corr = "unstructured",
  m = "m",
  data = long
)

# M2: independent domain RE
BBmm(
  y ~ x,
  random = ~ (0 + domain | id),
  dim = "domain",
  corr = "diag",
  m = "m",
  data = long
)
```

Inspect \(G=\Sigma\) with `fit$Sigma$id` and `fit$Corr$id`.  
M1 remains `random = ~ (1 | id)` (shared intercept).

Low-level stacking remains available via `multi_bb_stack()` for advanced use.

## Shared P-splines across dimensions

Additive `s()` terms are **shared** across outcomes (same \(f_j\) for every
\(\ell\)), consistent with the shared latent \(a_i\):

```{r multi-pspline}
set.seed(46)
n <- 70
id <- factor(seq_len(n))
x <- runif(n)
a <- rnorm(n, 0, 0.5)
f <- sin(2 * pi * x)
dat_s <- data.frame(
  id = id,
  x = x,
  y1 = rBB(n, 20L, 1 / (1 + exp(-(0.1 + f + a))), 0.18),
  y2 = rBB(n, 10L, 1 / (1 + exp(-(-0.2 + f + a))), 0.22)
)

fit_s <- BBmm(
  cbind(y1, y2) ~ s(x, ndx = 10, pord = 2),
  random = ~ (1 | id),
  m = c(20, 10),
  data = dat_s
)
fit_s$lambda
sum(fit_s$fhat[[1]])
```

Long format works the same way: `y ~ s(x)` with `dim = "domain"` (shared),
or domain-specific smooths:

```r
BBmm(
  y ~ hclus4 + s(wt, by = dim, k = 8) + s(fev1p, by = dim, k = 8),
  random = ~ (1 | id),
  dim = "dim",
  m = "m",
  data = long
)
```

COPD SGRQ port (BB-GAM `fit_pspline_M123` → this API):  
`scripts/fit_COPD_multi_pspline.R`.

# Notes

- Prefer `random = ~ (1 | id)` over legacy `random.formula = ~ id`.
- Weak RE / large \(\phi\) / few groups: Laplace ML may shrink \(\sigma\to 0\)
  while PROreg's h-likelihood keeps a small positive value.
- Additive P-splines (univariate or shared multivariate) are documented in the
  P-splines vignette (`plot_smooth()` for pointwise bands).
