---
title: "PROregTMB: benchmarks and joint-model results"
author: "Dae-Jin Lee"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PROregTMB: benchmarks and joint-model results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2,
  dpi = 120
)
```

# Overview

**PROregTMB** reimplements beta-binomial PRO regression with
[TMB](https://github.com/kaskr/adcomp) (automatic differentiation + Laplace
approximation). This vignette summarises numerical experiments comparing:

1. **`BBreg` / `BBmm`** against [PROreg](https://cran.r-project.org/package=PROreg)
   (including crossed multi-RE and multidimensional outcomes)
2. A **two-stage joint model (TSBB)** in the spirit of
   Galán-Arcicollar, Najera-Zuloaga & Lee (*SORT*, 2024)
3. A **one-stage joint model (`BBjm`)** with shared random effects, vs TSBB

Heavy simulations are **not** re-run here. Precomputed tables ship in
`inst/extdata/` (also produced by scripts under `scripts/`). Results below are
**pilots** (smaller \(N\) / \(n_{\mathrm{sim}}\) than the full paper grids).

```{r load-data}
extdata <- function(file) {
  # installed package
  p <- system.file("extdata", file, package = "PROregTMB")
  if (nzchar(p) && file.exists(p)) return(p)
  # knitr from vignettes/ or package root
  for (cand in c(
    file.path("inst", "extdata", file),
    file.path("..", "inst", "extdata", file),
    file.path("extdata", file)
  )) {
    if (file.exists(cand)) return(cand)
  }
  stop("Cannot find extdata file: ", file, call. = FALSE)
}

bench_re <- read.csv(extdata("bench_multi_RE.csv"))
tsbb_sum <- read.csv(extdata("tsbb_galan_summary.csv"))
tsbb_rep <- read.csv(extdata("tsbb_galan_reps.csv"))
bbjm_demo <- read.csv(extdata("demo_BBjm.csv"))
bbjm_sum <- read.csv(extdata("bbjm_vs_tsbb_summary.csv"))
bbjm_rep <- read.csv(extdata("bbjm_vs_tsbb_reps.csv"))
multi_sum <- read.csv(extdata("multi_BB_summary.csv"))
multi_rep <- read.csv(extdata("multi_BB_reps.csv"))
```

# Models in brief

| Function | Role |
|----------|------|
| `BBreg` | Marginal BB logistic regression (no RE) |
| `BBmm` | BB mixed model, Gaussian RE, Laplace |
| `BBjm` | One-stage BB + Weibull PH joint model |

**Longitudinal BB mixed model**

\[
y_{ij}\mid u_i \sim \mathrm{BB}(m,p_{ij},\phi),\quad
\operatorname{logit}(p_{ij}) = x_{ij}^\top\beta + z_{ij}^\top u_i.
\]

**One-stage joint model (`BBjm`)** — shared RI/RS, current-value association:

\[
\operatorname{logit}(p_{ij})=(\beta_0+u_{i0})+(\beta_1+u_{i1})t_{ij},
\]
\[
m_i(t)=m\cdot\operatorname{logit}^{-1}\big((\beta_0+u_{i0})+(\beta_1+u_{i1})t\big),
\]
\[
h_i(t)=w_1 w_2 t^{w_2-1}\exp\{\alpha\, m_i(t)\}.
\]

**Two-stage TSBB** fits `BBmm`, plugs \(\hat m_i(t)\) into `survival::coxph`.

# Multi random-effect timing (`BBmm`)

Crossed random effects `~ site + doc` (two variance components). Same data for
PROreg and PROregTMB; times are means of 3 replicates after TMB warm-up compile.

```{r bench-re-table}
tab_re <- data.frame(
  Scenario = bench_re$scenario,
  n = bench_re$n,
  n_RE = bench_re$n_re,
  PROreg_s = round(bench_re$time_proreg_mean, 3),
  TMB_s = round(bench_re$time_tmb_mean, 3),
  Speedup = round(bench_re$speedup, 2),
  cor_u = round(bench_re$cor_u, 4),
  sigma_site_P = round(bench_re$proreg_sigma_site, 3),
  sigma_site_T = round(bench_re$tmb_sigma_site, 3),
  sigma_doc_P = round(bench_re$proreg_sigma_doc, 3),
  sigma_doc_T = round(bench_re$tmb_sigma_doc, 3)
)
knitr::kable(tab_re, caption = "Multi-RE benchmark: timings and selected estimates.")
```

```{r bench-re-plot, fig.cap="Wall-clock time for BBmm with two RE components."}
op <- par(mar = c(5, 4, 2, 1))
barplot(
  rbind(bench_re$time_proreg_mean, bench_re$time_tmb_mean),
  beside = TRUE,
  names.arg = bench_re$scenario,
  legend.text = c("PROreg", "PROregTMB"),
  args.legend = list(x = "topleft", bty = "n"),
  ylab = "Mean elapsed time (s)",
  col = c("gray70", "steelblue")
)
par(op)
```

**Takeaway.** Estimates agree closely (`cor(û) ≥ 0.998`). Speedup grows with
problem size, from about **3×** (n = 150) to **17×** (n = 1920, 280 RE).

# Two-stage joint model pilot (TSBB)

Design follows Galán-Arcicollar et al. (*SORT* 2024, Sec. 6): visits at
\(0,1,2,5\); Weibull baseline \((w_1,w_2)=(0.1,1.6)\); random intercept and
slope; association parameter \(\alpha\).

**Pilot settings (not the full paper grid):** \(N=80\), \(n_{\mathrm{sim}}=8\),
four representative \((\phi,\alpha)\) cells. Stage-1 engine is either
`PROreg::BBmm` or `PROregTMB::BBmm`; stage-2 `coxph` is identical.

```{r tsbb-table}
tsbb_view <- within(tsbb_sum, {
  pct_bias <- round(pct_bias, 3)
  ESD <- round(ESD, 4)
  CP95 <- round(CP95, 1)
  time_mean <- round(time_mean, 2)
  speedup <- round(speedup, 1)
  mean_alpha <- round(mean_alpha, 4)
})
knitr::kable(
  tsbb_view[, c("tag", "engine", "alpha_true", "mean_alpha", "pct_bias",
                "ESD", "CP95", "pct_conv", "time_mean", "speedup")],
  col.names = c("Cell", "Engine", "α", "Mean α̂", "%Bias", "ESD", "CP95",
                "%Conv", "Time (s)", "Speedup"),
  caption = "TSBB pilot: association parameter α."
)
```

```{r tsbb-time-plot, fig.cap="TSBB full-pipeline time by cell and engine."}
cells <- unique(tsbb_sum$tag)
P <- tsbb_sum$time_mean[tsbb_sum$engine == "PROreg"]
Tm <- tsbb_sum$time_mean[tsbb_sum$engine == "PROregTMB"]
op <- par(mar = c(8, 4, 2, 1))
barplot(
  rbind(P, Tm),
  beside = TRUE,
  names.arg = cells,
  las = 2,
  legend.text = c("PROreg", "PROregTMB"),
  args.legend = list(x = "topright", bty = "n"),
  ylab = "Mean time (s)",
  col = c("gray70", "steelblue"),
  cex.names = 0.75
)
par(op)
```

```{r tsbb-alpha-scatter, fig.cap="Replicate-level agreement of α̂ (PROreg vs PROregTMB)."}
ok <- complete.cases(tsbb_rep$alpha_proreg, tsbb_rep$alpha_tmb)
plot(tsbb_rep$alpha_proreg[ok], tsbb_rep$alpha_tmb[ok],
     xlab = expression(hat(alpha)[PROreg]),
     ylab = expression(hat(alpha)[TMB]),
     pch = 19, col = adjustcolor("steelblue", 0.7))
abline(0, 1, lty = 2)
legend("topleft", bty = "n",
       legend = sprintf("cor = %.3f", cor(tsbb_rep$alpha_proreg[ok],
                                          tsbb_rep$alpha_tmb[ok])))
```

**Takeaway.** On this pilot, PROregTMB recovers essentially the same \(\hat\alpha\)
as PROreg (replicate correlation \(\approx 0.99\)) with about **30–47×** lower
wall-clock time. Coverage and ESD are comparable; `%Bias` for very small
\(\alpha\) is noisy at \(n_{\mathrm{sim}}=8\).

Reproduce / extend:

```r
# quick pilot
# Rscript scripts/sim_tsbb_galan.R --quick --nsim=8 --N=80
# full paper grid (slow with PROreg):
# Rscript scripts/sim_tsbb_galan.R --nsim=200 --N=250
```

# One-stage joint model (`BBjm`) vs TSBB

## Demo replicate

Single illustrative replicate (Scenario-1-like: \(N=120\), \(m=24\),
\(\phi=0.5\), \(\alpha=0.05\)).

```{r bbjm-table}
knitr::kable(
  within(bbjm_demo, {
    true <- round(true, 4)
    TSBB <- round(TSBB, 4)
    BBjm <- round(BBjm, 4)
    diff_jm_ts <- round(diff_jm_ts, 4)
  }),
  caption = "One replicate: TSBB (PROregTMB::BBmm + coxph) vs BBjm."
)
```

In this run, **`BBjm` recovered \(\hat\alpha \approx 0.051\)** (truth \(0.05\))
versus TSBB \(\hat\alpha \approx 0.042\), and also estimated the Weibull
baseline \((w_1,w_2)\). See `scripts/demo_BBjm.R`.

```{r bbjm-usage, eval=FALSE}
library(PROregTMB)
# long: id, time, y
# surv: id, time, status (1 = event)
fit <- BBjm(long, surv, m = 24)
summary(fit)
```

## Monte Carlo pilot (`BBjm` vs TSBB)

Same SORT 2024 design; stage-1 of TSBB uses **`PROregTMB::BBmm`** (so the
comparison isolates one-stage vs two-stage, not the BB engine).

**Pilot settings:** \(N=100\), \(n_{\mathrm{sim}}=40\), five \((\phi,\alpha)\)
cells (Scenario 1 and 2). Script: `scripts/sim_BBjm_vs_TSBB.R`.

```{r bbjm-mc-table}
jm_view <- within(bbjm_sum, {
  pct_bias <- round(pct_bias, 3)
  RMSE <- round(RMSE, 4)
  ESD <- round(ESD, 4)
  CP95 <- round(CP95, 1)
  time_mean <- round(time_mean, 2)
  mean_alpha <- round(mean_alpha, 4)
})
knitr::kable(
  jm_view[, c("tag", "method", "alpha_true", "mean_alpha", "pct_bias",
              "RMSE", "ESD", "CP95", "pct_conv", "time_mean")],
  col.names = c("Cell", "Method", "α", "Mean α̂", "%Bias", "RMSE", "ESD",
                "CP95", "%Conv", "Time (s)"),
  caption = "BBjm vs TSBB pilot: association parameter α."
)
```

```{r bbjm-mc-time, fig.cap="Mean fit time by cell (TSBB uses TMB stage-1)."}
cells <- unique(bbjm_sum$tag)
ts <- bbjm_sum$time_mean[bbjm_sum$method == "TSBB"]
jm <- bbjm_sum$time_mean[bbjm_sum$method == "BBjm"]
op <- par(mar = c(8, 4, 2, 1))
barplot(
  rbind(ts, jm),
  beside = TRUE,
  names.arg = cells,
  las = 2,
  legend.text = c("TSBB (TMB)", "BBjm"),
  args.legend = list(x = "topright", bty = "n"),
  ylab = "Mean time (s)",
  col = c("gray70", "darkorange"),
  cex.names = 0.7
)
par(op)
```

```{r bbjm-mc-alpha, fig.cap="Replicate-level α̂: TSBB vs BBjm."}
ok <- complete.cases(bbjm_rep$alpha_tsbb, bbjm_rep$alpha_bbjm)
plot(bbjm_rep$alpha_tsbb[ok], bbjm_rep$alpha_bbjm[ok],
     xlab = expression(hat(alpha)[TSBB]),
     ylab = expression(hat(alpha)[BBjm]),
     pch = 19, col = adjustcolor("darkorange", 0.55))
abline(0, 1, lty = 2)
legend("topleft", bty = "n",
       legend = sprintf("cor = %.3f",
                        cor(bbjm_rep$alpha_tsbb[ok], bbjm_rep$alpha_bbjm[ok])))
```

**Takeaway.** Both methods converged in **100%** of replicates. Bias, RMSE and
95% CP for \(\alpha\) are **similar** across cells; neither dominates on this
pilot. With TMB already in stage-1, **TSBB is faster** (~0.2–0.3 s vs ~1.1–1.3 s
for `BBjm`). One-stage estimation still returns Weibull baseline parameters in
a single fit. Stronger claims need the full paper grid
(\(N=250\), \(n_{\mathrm{sim}}\ge 100\)).

Reproduce / extend:

```r
# Rscript scripts/sim_BBjm_vs_TSBB.R --quick --nsim=40 --N=100
# Rscript scripts/sim_BBjm_vs_TSBB.R --nsim=100 --N=250
```

# Multidimensional BBMM (Najera-Zuloaga et al., 2024)


Shared random-effects joint analysis of \(L=3\) BB outcomes (Statistical Modelling;
Table 1):

\[
Y_{il}\mid u_i \sim \mathrm{BB}(m_l,p_{il},\phi_l),\quad
\operatorname{logit}(p_{il})=\gamma_{0l}+\gamma_{1l}x_i+u_i,\quad
u_i\sim N(0,\sigma_u^2).
\]

Truth: \((\gamma_{0},\gamma_{1},\phi,m)\) =
\((1,-1.5,0.5,20)\), \((-2,2.5,1,10)\), \((0.5,0.5,1.5,4)\);
\(x\sim U(0,1)\); \(n=100\). Pilot: \(n_{\mathrm{sim}}=30\), \(\sigma_u\in\{0.5,1\}\).

```{r multi-slopes}
sl <- multi_sum[multi_sum$param %in% c("g1_1", "g1_2", "g1_3", "sigma"), ]
sl_view <- within(sl, {
  bias <- round(bias, 3)
  RMSE <- round(RMSE, 3)
  ESD <- round(ESD, 3)
  CP95 <- round(CP95, 1)
  time_mean <- round(time_mean, 2)
})
knitr::kable(
  sl_view[, c("sigma_u", "engine", "param", "true", "bias", "RMSE", "ESD",
              "CP95", "time_mean")],
  caption = "Multidimensional BBMM: slopes and shared sigma (PROreg vs TMB)."
)
```

```{r multi-time, fig.cap="Multidimensional BBmm wall-clock time."}
agg <- aggregate(cbind(time_P, time_T) ~ sigma_u, multi_rep, mean)
op <- par(mar = c(4, 4, 2, 1))
barplot(
  rbind(agg$time_P, agg$time_T),
  beside = TRUE,
  names.arg = paste0("sigma=", agg$sigma_u),
  legend.text = c("PROreg", "PROregTMB"),
  args.legend = list(x = "topright", bty = "n"),
  ylab = "Mean time (s)",
  col = c("gray70", "steelblue")
)
par(op)
```

```{r multi-cor}
ok <- complete.cases(multi_rep$g1_1_P, multi_rep$g1_1_T)
cat(sprintf(
  "Replicate correlation of slope g1 (dim 1): %.4f\n",
  cor(multi_rep$g1_1_P[ok], multi_rep$g1_1_T[ok])
))
cat(sprintf(
  "Mean speedup: %.1fx\n",
  mean(multi_rep$time_P) / mean(multi_rep$time_T)
))
```

**Takeaway.** Both engines recover slopes with similar bias/RMSE/CP; TMB is about
**30×** faster. Dimension 3 (\(m=4\)) is noisier for both when \(\sigma_u=1\)
(larger ESD), matching the paper’s caution about extreme shared-RE scenarios.

Reproduce:

```r
# Rscript scripts/sim_multi_BB.R --quick --nsim=30
# full: Rscript scripts/sim_multi_BB.R --nsim=100  # sigma in {0.5,1,1.5}
```

# Session info

```{r session}
sessionInfo()
```
