---
title: "Getting started with PROregTMB"
author: "Dae-Jin Lee"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with PROregTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 4,
  dpi = 120
)
```

# Overview

**PROregTMB** fits beta-binomial (BB) regression models for patient-reported
outcomes (bounded integer scores) using
[TMB](https://github.com/kaskr/adcomp) for automatic differentiation and
Laplace approximation.

| Function | Role |
|----------|------|
| `dBB` / `rBB` | BB density and simulation |
| `BBest` | Moment / MLE estimate of dispersion \(\phi\) |
| `BBreg` | Marginal BB logistic regression |
| `BBmm` | BB mixed model (Gaussian RE + Laplace) |
| `BBjm` | One-stage BB + Weibull joint model |
| `s()` | Additive P-splines (Eilers–Marx) inside `BBreg` / `BBmm` |

Related vignettes: mixed / multivariate models; additive P-splines;
benchmarks and joint-model pilots.

```{r lib}
library(PROregTMB)
```

The first call to `BBreg()` / `BBmm()` / `BBjm()` compiles a TMB template
(requires a C++ toolchain). A short status line is printed; suppress it with
`options(PROregTMB.verbose_compile = FALSE)`. Later calls reuse the DLL.

# Beta-binomial building blocks

A score \(y\in\{0,\ldots,m\}\) follows
\(y\sim\mathrm{BB}(m,p,\phi)\) with mean \(mp\) and overdispersion \(\phi>0\).

```{r bb-sim}
set.seed(1)
m <- 10L
y <- rBB(k = 8, m = m, p = 0.4, phi = 0.25)
y
# PMF on 0, ..., m
round(dBB(m = m, p = 0.4, phi = 0.25), 4)
```

`BBest()` estimates \(\phi\) from a vector of scores (method of moments or MLE).

```{r bbest}
set.seed(2)
y <- rBB(200, m = 10, p = 0.55, phi = 0.3)
BBest(y, m = 10)
```

# Marginal BB logistic regression (`BBreg`)

Model:

\[
y_i\sim\mathrm{BB}(m_i,p_i,\phi),\qquad
\operatorname{logit}(p_i)=x_i^\top\beta.
\]

```{r bbreg}
set.seed(18)
n <- 400
m <- 10L
x <- rnorm(n, 5, 3)
p <- 1 / (1 + exp(-(-10 + 2 * x)))
y <- rBB(n, m, p, phi = 1.2)
dat <- data.frame(y, x)

fit <- BBreg(y ~ x, m = m, data = dat)
fit
```

Primary fields follow the formulation (`$beta`, `$phi`, `$p`); PROreg-style
aliases (`$coefficients`, `$fitted.values`) remain available.

```{r bbreg-fields}
fit$beta
fit$phi
head(fit$p)
summary(fit)
```

# Formulation fields

```{r model-lines}
fit$model
```

# Next steps

- Random intercepts / slopes and multivariate outcomes: vignette
  **PROregTMB mixed and multivariate models**.
- Smooth effects with P-splines and Marra–Wood (2012) bands: vignette
  **Additive P-splines in PROregTMB**.
- Simulation pilots vs PROreg / TSBB / `BBjm`: vignette
  **PROregTMB benchmarks and joint-model results**.
