---
title: "PROregTMB: benchmarks and joint-model results"
author: "Dae-Jin Lee"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PROregTMB: benchmarks and joint-model results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2,
  dpi = 120
)
```

# Overview

**PROregTMB** reimplements beta-binomial PRO regression with
[TMB](https://github.com/kaskr/adcomp) (automatic differentiation + Laplace
approximation). This vignette summarises numerical experiments comparing:

1. **`BBreg` / `BBmm`** against [PROreg](https://cran.r-project.org/package=PROreg)
2. A **two-stage joint model (TSBB)** in the spirit of
   Galán-Arcicollar, Najera-Zuloaga & Lee (*SORT*, 2024)
3. A **one-stage joint model (`BBjm`)** with shared random effects

Heavy simulations are **not** re-run here. Precomputed tables ship in
`inst/extdata/` (also produced by scripts under `scripts/`).

```{r load-data}
extdata <- function(file) {
  # installed package
  p <- system.file("extdata", file, package = "PROregTMB")
  if (nzchar(p) && file.exists(p)) return(p)
  # knitr from vignettes/ or package root
  for (cand in c(
    file.path("inst", "extdata", file),
    file.path("..", "inst", "extdata", file),
    file.path("extdata", file)
  )) {
    if (file.exists(cand)) return(cand)
  }
  stop("Cannot find extdata file: ", file, call. = FALSE)
}

bench_re <- read.csv(extdata("bench_multi_RE.csv"))
tsbb_sum <- read.csv(extdata("tsbb_galan_summary.csv"))
tsbb_rep <- read.csv(extdata("tsbb_galan_reps.csv"))
bbjm_demo <- read.csv(extdata("demo_BBjm.csv"))
```

# Models in brief

| Function | Role |
|----------|------|
| `BBreg` | Marginal BB logistic regression (no RE) |
| `BBmm` | BB mixed model, Gaussian RE, Laplace |
| `BBjm` | One-stage BB + Weibull PH joint model |

**Longitudinal BB mixed model**

\[
y_{ij}\mid u_i \sim \mathrm{BB}(m,p_{ij},\phi),\quad
\operatorname{logit}(p_{ij}) = x_{ij}^\top\beta + z_{ij}^\top u_i.
\]

**One-stage joint model (`BBjm`)** — shared RI/RS, current-value association:

\[
\operatorname{logit}(p_{ij})=(\beta_0+u_{i0})+(\beta_1+u_{i1})t_{ij},
\]
\[
m_i(t)=m\cdot\operatorname{logit}^{-1}\big((\beta_0+u_{i0})+(\beta_1+u_{i1})t\big),
\]
\[
h_i(t)=w_1 w_2 t^{w_2-1}\exp\{\alpha\, m_i(t)\}.
\]

**Two-stage TSBB** fits `BBmm`, plugs \(\hat m_i(t)\) into `survival::coxph`.

# Multi random-effect timing (`BBmm`)

Crossed random effects `~ site + doc` (two variance components). Same data for
PROreg and PROregTMB; times are means of 3 replicates after TMB warm-up compile.

```{r bench-re-table}
tab_re <- data.frame(
  Scenario = bench_re$scenario,
  n = bench_re$n,
  n_RE = bench_re$n_re,
  PROreg_s = round(bench_re$time_proreg_mean, 3),
  TMB_s = round(bench_re$time_tmb_mean, 3),
  Speedup = round(bench_re$speedup, 2),
  cor_u = round(bench_re$cor_u, 4),
  sigma_site_P = round(bench_re$proreg_sigma_site, 3),
  sigma_site_T = round(bench_re$tmb_sigma_site, 3),
  sigma_doc_P = round(bench_re$proreg_sigma_doc, 3),
  sigma_doc_T = round(bench_re$tmb_sigma_doc, 3)
)
knitr::kable(tab_re, caption = "Multi-RE benchmark: timings and selected estimates.")
```

```{r bench-re-plot, fig.cap="Wall-clock time for BBmm with two RE components."}
op <- par(mar = c(5, 4, 2, 1))
barplot(
  rbind(bench_re$time_proreg_mean, bench_re$time_tmb_mean),
  beside = TRUE,
  names.arg = bench_re$scenario,
  legend.text = c("PROreg", "PROregTMB"),
  args.legend = list(x = "topleft", bty = "n"),
  ylab = "Mean elapsed time (s)",
  col = c("gray70", "steelblue")
)
par(op)
```

**Takeaway.** Estimates agree closely (`cor(û) ≥ 0.998`). Speedup grows with
problem size, from about **3×** (n = 150) to **17×** (n = 1920, 280 RE).

# Two-stage joint model pilot (TSBB)

Design follows Galán-Arcicollar et al. (*SORT* 2024, Sec. 6): visits at
\(0,1,2,5\); Weibull baseline \((w_1,w_2)=(0.1,1.6)\); random intercept and
slope; association parameter \(\alpha\).

**Pilot settings (not the full paper grid):** \(N=80\), \(n_{\mathrm{sim}}=8\),
four representative \((\phi,\alpha)\) cells. Stage-1 engine is either
`PROreg::BBmm` or `PROregTMB::BBmm`; stage-2 `coxph` is identical.

```{r tsbb-table}
tsbb_view <- within(tsbb_sum, {
  pct_bias <- round(pct_bias, 3)
  ESD <- round(ESD, 4)
  CP95 <- round(CP95, 1)
  time_mean <- round(time_mean, 2)
  speedup <- round(speedup, 1)
  mean_alpha <- round(mean_alpha, 4)
})
knitr::kable(
  tsbb_view[, c("tag", "engine", "alpha_true", "mean_alpha", "pct_bias",
                "ESD", "CP95", "pct_conv", "time_mean", "speedup")],
  col.names = c("Cell", "Engine", "α", "Mean α̂", "%Bias", "ESD", "CP95",
                "%Conv", "Time (s)", "Speedup"),
  caption = "TSBB pilot: association parameter α."
)
```

```{r tsbb-time-plot, fig.cap="TSBB full-pipeline time by cell and engine."}
cells <- unique(tsbb_sum$tag)
P <- tsbb_sum$time_mean[tsbb_sum$engine == "PROreg"]
Tm <- tsbb_sum$time_mean[tsbb_sum$engine == "PROregTMB"]
op <- par(mar = c(8, 4, 2, 1))
barplot(
  rbind(P, Tm),
  beside = TRUE,
  names.arg = cells,
  las = 2,
  legend.text = c("PROreg", "PROregTMB"),
  args.legend = list(x = "topright", bty = "n"),
  ylab = "Mean time (s)",
  col = c("gray70", "steelblue"),
  cex.names = 0.75
)
par(op)
```

```{r tsbb-alpha-scatter, fig.cap="Replicate-level agreement of α̂ (PROreg vs PROregTMB)."}
ok <- complete.cases(tsbb_rep$alpha_proreg, tsbb_rep$alpha_tmb)
plot(tsbb_rep$alpha_proreg[ok], tsbb_rep$alpha_tmb[ok],
     xlab = expression(hat(alpha)[PROreg]),
     ylab = expression(hat(alpha)[TMB]),
     pch = 19, col = adjustcolor("steelblue", 0.7))
abline(0, 1, lty = 2)
legend("topleft", bty = "n",
       legend = sprintf("cor = %.3f", cor(tsbb_rep$alpha_proreg[ok],
                                          tsbb_rep$alpha_tmb[ok])))
```

**Takeaway.** On this pilot, PROregTMB recovers essentially the same \(\hat\alpha\)
as PROreg (replicate correlation \(\approx 0.99\)) with about **30–47×** lower
wall-clock time. Coverage and ESD are comparable; `%Bias` for very small
\(\alpha\) is noisy at \(n_{\mathrm{sim}}=8\).

Reproduce / extend:

```r
# quick pilot
# Rscript scripts/sim_tsbb_galan.R --quick --nsim=8 --N=80
# full paper grid (slow with PROreg):
# Rscript scripts/sim_tsbb_galan.R --nsim=200 --N=250
```

# One-stage joint model (`BBjm`) vs TSBB

Single illustrative replicate (Scenario-1-like: \(N=120\), \(m=24\),
\(\phi=0.5\), \(\alpha=0.05\)).

```{r bbjm-table}
knitr::kable(
  within(bbjm_demo, {
    true <- round(true, 4)
    TSBB <- round(TSBB, 4)
    BBjm <- round(BBjm, 4)
    diff_jm_ts <- round(diff_jm_ts, 4)
  }),
  caption = "One replicate: TSBB (PROregTMB::BBmm + coxph) vs BBjm."
)
```

In this run, **`BBjm` recovered \(\hat\alpha \approx 0.051\)** (truth \(0.05\))
versus TSBB \(\hat\alpha \approx 0.042\), and also estimated the Weibull
baseline \((w_1,w_2)\). Wall-clock was about **1.4 s** for `BBjm` vs
**~13 s** for TSBB on the same machine (see `scripts/demo_BBjm.R`).

```{r bbjm-usage, eval=FALSE}
library(PROregTMB)
# long: id, time, y
# surv: id, time, status (1 = event)
fit <- BBjm(long, surv, m = 24)
summary(fit)
```

# How to argue that `BBjm` is better

A fair claim needs a Monte Carlo study (same truth as Galán-Arcicollar et al.):

| Metric for \(\alpha\) | Prefer `BBjm` if… |
|-----------------------|-------------------|
| %Bias / RMSE | smaller than TSBB |
| 95% CP | closer to nominal |
| % convergence | higher / more stable |
| Time | secondary |

`BBjm` is expected to reduce **plug-in bias** of two-stage methods; TSBB may
remain competitive on speed when stage-1 uses TMB. Report the full
\((\phi,\alpha)\) grid with large \(n_{\mathrm{sim}}\) before strong conclusions.

# Session info

```{r session}
sessionInfo()
```
